Inferring the existence and direction of causal associations in the face of measurement error

### 2017-02-09 docs/manuscript.Rmd
updated

Inference of the causal structure that induces correlations between two traits can be achieved by combining genetic associations with a mediation-based approach, as is done in the causal inference test (CIT) and others. However, we show that measurement error in the phenotypes can lead to mediation-based approaches inferring the wrong causal direction, and that increasing sample sizes has the adverse effect of increasing confidence in the wrong answer. Here we introduce an extension to Mendelian randomisation, a method that uses genetic associations in an instrumentation framework, that enables inference of the causal direction between traits, with some advantages. First, it is less susceptible to bias in the presence of measurement error; second, it is more statistically efficient; third, it can be performed using only summary level data from genome-wide association studies; and fourth, its sensitivity to measurement error can be evaluated. We apply the method to infer the causal direction between DNA methylation and gene expression levels. Our results demonstrate that, in general, DNA methylation is more likely to be the causal factor, but this result is highly susceptible to bias induced by systematic differences in measurement error between the platforms. We emphasise that, where possible, implementing MR alongside other approaches such as CIT is important to triangulate reliable conclusions about causality.

@@

Measurement error will of course arise in other types of data too, for example when measuring BMI one is typically interested in using this as a proxy for adiposity, but it is clear that the correlation between BMI and underlying adiposity is not perfect [@Ahima2013]. A similar problem of biological misspecification is unavoidable in disease diagnosis, and measuring behaviour such as smoking or diet is notoriously difficult to do accurately. Measurement error can also be introduced after the data has been collected, for example the transformation of non-normal data for the purpose of statistical analysis will lead to a new variable that will typically incur both changes in scale and imprecision (noise) compared to the original variable. The sources of measurement error are not limited to this list [@Hernan2009], and its impact has been explored in the epidemiological literature extensively [@LeCessie2012; @Blakely2013]. Given the near-ubiquitous presence of measurement error in phenomic data it is vital to understand its impact on the tools we use for causal inference.

@@

By utilising genetic instruments in different ways, mediation-based analysis and MR models have properties that confer some advantages and some disadvantages for reliable causal inference. An important advantage of mediation-based analysis is that in order to infer the causal direction the true underlying biology of the genetic instrument doesn't necessarily need to be known as long as it associates with the putative exposure. In the CIT framework (described fully in the Methods) for example, the test statistic is different if you test for the exposure causing the outcome or the outcome causing the exposure, allowing the researcher to infer the direction of causality between two variables by performing the test in both directions and choosing the model with the strongest evidence. The CIT also has the valuable property of being able to distinguish between several putative causal graphs that link the traits with the SNP (`r cfig("dag_figure")`). Such is not the case for MR, where in order to infer the direction of causality between two traits the instrument must be known to influence the exposure primarily, associating with the outcome only through the exposure.

@@

MR does however have some important advantages over the mediation-based approaches. The mediation-based approaches require that the exposure, outcome and instrumental variables are all measured in the same data, whereas recent extensions to MR circumvent this requirement, allowing causal inference to be drawn when exposure variables and outcome variables are measured in different samples [@Pierce2013]. This has the crucial advantage of improving statistical power by allowing analysis in much larger sample sizes, and dramatically expands the breadth of possible phenotypic relationships that can be evaluated [@Hemani2016]. Additionally, when MR assumptions are satisfied the method is substantially more robust to there being measurement error in the exposure variable. Indeed instrumental variable (IV) analysis was in part initially introduced as a correction for measurement error in the exposure [@Ashenfelter1994], whereas it has been noted that both classic mediation-based analyses [@LeCessie2012; @Nagarajan2013; @Shpitser2010; @Blakely2013] and mediation-based methods that use instrumental variables [@Wang2015; @Lagani2015] are prone to be unreliable in its presence.

@@

>>>>

Inference of the causal structure that induces correlations between two traits can be achieved by combining genetic associations with a mediation-based approach, as is done in the causal inference test (CIT) and others. However, we show that measurement error in the phenotypes can lead to mediation-based approaches inferring the wrong causal direction, and that increasing sample sizes has the adverse effect of increasing confidence in the wrong answer. Here we introduce an extension to Mendelian randomisation, a method that uses genetic associations in an instrumentation framework, that enables inference of the causal direction between traits, with some advantages. First, it is less susceptible to bias in the presence of measurement error; second, it is more statistically efficient; third, it can be performed using only summary level data from genome-wide association studies; and fourth, its sensitivity to measurement error can be evaluated. We apply the method to infer the causal direction between DNA methylation and gene expression levels. Our results demonstrate that, in general, DNA methylation is more likely to be the causal factor, but this result is highly susceptible to bias induced by systematic differences in measurement error between the platforms. We emphasise that, where possible, implementing MR and appropriate sensitivity analyses alongside other approaches such as CIT is important to triangulate reliable conclusions about causality.

@@

Measurement error will of course arise in other types of data too. For example when measuring BMI one is typically interested in using this as a proxy for adiposity, but it is clear that the correlation between BMI and underlying adiposity is not perfect [@Ahima2013]. A similar problem of biological misspecification is unavoidable in disease diagnosis, and measuring behaviour such as smoking or diet is notoriously difficult to do accurately. Measurement error can also be introduced after the data has been collected, for example the transformation of non-normal data for the purpose of statistical analysis will lead to a new variable that will typically incur both changes in scale and imprecision (noise) compared to the original variable. The sources of measurement error are not limited to this list [@Hernan2009], and its impact has been explored in the epidemiological literature extensively [@LeCessie2012; @Blakely2013]. Given the near-ubiquitous presence of measurement error in phenomic data it is vital to understand its impact on the tools we use for causal inference.

@@

By utilising genetic instruments in different ways, mediation-based analysis and MR models have properties that confer some advantages and some disadvantages for reliable causal inference. In the CIT framework (described fully in the Methods) for example, the test statistic is different if you test for the exposure causing the outcome or the outcome causing the exposure, allowing the researcher to infer the direction of causality between two variables by performing the test in both directions and choosing the model with the strongest evidence. The CIT also has the valuable property of being able to distinguish between several putative causal graphs that link the traits with the SNP (`r cfig("dag_figure")`). Such is not the case for MR, where in order to infer the direction of causality between two traits the instrument must be known to influence the exposure primarily, associating with the outcome only through the exposure.

@@

MR has some important advantages over the mediation-based approaches. The mediation-based approaches require that the exposure, outcome and instrumental variables are all measured in the same data, whereas recent extensions to MR circumvent this requirement, allowing causal inference to be drawn when exposure variables and outcome variables are measured in different samples [@Pierce2013]. This has the crucial advantage of improving statistical power by allowing analysis in much larger sample sizes, and dramatically expands the breadth of possible phenotypic relationships that can be evaluated [@Hemani2016]. Additionally, when MR assumptions are satisfied the method is robust to there being measurement error in the exposure variable. Indeed instrumental variable (IV) analysis was in part initially introduced as a correction for measurement error in the exposure [@Ashenfelter1994], whereas it has been noted that both classic mediation-based analyses [@LeCessie2012; @Nagarajan2013; @Shpitser2010; @Blakely2013] and mediation-based methods that use instrumental variables [@Wang2015; @Lagani2015] are prone to be unreliable in its presence.

@@

## Model
